Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-05 paper-correlations.tex

\begin{figure*}[pt]
  \centering
  \subfigure[Ceramic 1]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic1.png}}
  \hfill
  \subfigure[Ceramic 2]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic2.png}}
  \hfill
  \subfigure[Ceramic 3]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic3.png}}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/s2-xct.png}
  \includegraphics[width=0.3\linewidth]{samples/c2-xct.png}
  \includegraphics[width=0.3\linewidth]{samples/l2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/surf2-xct.png}
  \includegraphics[width=0.3\linewidth]{samples/surfvoid-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/pore-size-xct.png}
  \includegraphics[width=0.3\linewidth]{samples/chord-length-xct.png}
  \caption[]{Visualization of XCT images and plots of correlation functions for
    these images.}
  \label{fig:real-xct}
\end{figure*}

The \cref{fig:real-sem} presents three 2D images of sandstones and carbonate with computed correlation functions. In this example,
unlike for quite isotropic ceramics, we evaluated CFs in the directional manner. The $S_2$ graph immediately provides us
with porosity values. It also highlights severe anisotropy in sandstone SEM images. For 2D images numerous patches of pores are
not connected to each other and, thus, $S_2$ and $C_2$ are very different. Cluster functions show how connectivity decays with
length, also we can easily observe if given pore structure percolates in a given direction - none of the images percolates
(we highlighted the clusters to make this statement easily verifiable with the naked eye). Note that $C_2$ and $L_2$ were computed
with correlations lengths equal to image dimensions explore connectivity - for other CFs this leads to noised curves due to smaller number of sampled points (for example, see correlation functions' graphs in \cite{Euras2012}).
Lineal paths are close for sandstones (with lower $L_2$ values for Sandstone 1 where large pores are occluded with clays or grain fragments),
but for carbonate with much smaller apertures of pores it is way below $L_2$ for sandstones. Also, $L_2$ suggests that computations
with correlations lengths beyond where lineal path goes to zero are not very informative. For very irregular pores on 2D SEM images,
$P$ and $p$ functions look much less similar compared to the ceramics case on \cref{fig:real-xct}. For 2D images it is also visually
easy to analyze full CF maps, as presented on \cref{fig:real-sem_maps} for Sandstone 1 sample. Full correlation maps show expanded version of directional functions -
asymmetry is immediately evidenced from $S_2$, $F_ss$ and $F_sv$ maps. Moreover, the map for $C_2$ shows higher connectivity in y-direction
as compared to x-axis. The interpretation of full correlation maps is less straightforward, but anisotropy patterns in directions different from
those presented in \cref{fig:directions} can be spotted.

\begin{figure*}[pt]
  \centering
  \subfigure[Sandstone 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone1.png}}
  \hfill
  \subfigure[Sandstone 2]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone2.png}}
  \hfill
  \subfigure[Carbonate 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/carbonate1.png}}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/s2-sem.png}
  \includegraphics[width=0.3\linewidth]{samples/c2-sem.png}
  \includegraphics[width=0.3\linewidth]{samples/l2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/surf2-sem.png}
  \includegraphics[width=0.3\linewidth]{samples/surfvoid-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{samples/pore-size-sem.png}
  \includegraphics[width=0.3\linewidth]{samples/chord-length-sem.png}
  \caption[]{SEM images and plots of correlation functions for these images.}
  \label{fig:real-sem}
\end{figure*}

Finally, we conclude this section with the example of multi-phase soil composite for which we computed cross-correlation functions \cref{fig:real-multiphase}.
The graph on the right side of the figure shows cross-correlations between different pairs of phases (normalized by each phase's ratio). One can easily conclude
that organo-mineral complex (phase 2) is mainly located on pore (phase 1) walls. Heavy minerals (phase 4) do not cross-correlate with pores, as
they are mainly lie within the mineral solid phase (phase 3) - they do show high cross-correlations. Other CFs can be easily computed for each separate phase,
while $F_sv$ can potentially be extended beyond void (or pores) and be represented by any other phase or their superposition.

@@

\section{Discussion and outlook}
In the light of the end of the previous section it is important to clarify that available
(video) RAM or speed of computations are not the only considerations behind choosing between full correlation map
and directional linear scans. Directional CFs are extremely useful in describing or reconstructing
anisotropic structures \cite{jiao2014chawla,EPL1,tensorPRE}. At the same time, they also much easier to
work with and good candidates for data compression and feature extraction \cite{KarsaninaEJSS}. On the other hand,
full map for $S_2$ allows computing ensemble averaged two-point probability - in the similar fashion it is
measured with scattering experiments. The usefulness of full correlation maps for other CFs is questionable,
as they are either non-measurable experimentally, e.g., \cite{Gommes_notchord}, or non-applicable for stochastic reconstructions.
The latter refers to our ability to reconstruct almost any structure from its $S_2$ map \cite{fullwood2008}. And while we
have fast and efficient techniques such as phase-recovery to reconstruct the structure from its $S_2$ correlation map, the utilization
of other CFs is complicated \cite{cherkasov2021,cherkasovIC}. Simulated-annealing based methods that operate on the full map are, 
to the best of our knowledge, were not developed to date. This is the reason we did not invest into improvement on existing approaches to compute maps for $L_2$.

>>>>

\begin{figure*}[pt]
  \centering
  \subfigure[Ceramic 1]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic1.png}}
  \hfill
  \subfigure[Ceramic 2]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic2.png}}
  \hfill
  \subfigure[Ceramic 3]{
    \includegraphics[width=0.3\linewidth]{samples/Ceramic3.png}}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/s2-xct.png}
  \includegraphics[width=0.32\linewidth]{samples/c2-xct.png}
  \includegraphics[width=0.32\linewidth]{samples/l2-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/surf2-xct.png}
  \includegraphics[width=0.32\linewidth]{samples/surfvoid-xct.png}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/pore-size-xct.png}
  \includegraphics[width=0.32\linewidth]{samples/chord-length-xct.png}
  \caption[]{Visualization of XCT images and plots of correlation functions for
    these images.}
  \label{fig:real-xct}
\end{figure*}

The \cref{fig:real-sem} presents three 2D images of sandstones and carbonate with computed correlation functions. In this example,
unlike for quite isotropic ceramics, we evaluated CFs in the directional manner. The $S_2$ graph immediately provides us
with porosity values. It also highlights severe anisotropy in sandstone SEM images. For 2D images numerous patches of pores are
not connected to each other and, thus, $S_2$ and $C_2$ are very different. Cluster functions show how connectivity decays with
length, also we can easily observe if given pore structure percolates in a given direction - none of the images percolates
(we highlighted the clusters to make this statement easily verifiable with the naked eye). Note that $C_2$ and $L_2$ were computed
with correlations lengths equal to image dimensions explore connectivity - for other CFs this leads to noised curves due to smaller number of sampled points (for example, see correlation functions' graphs in \cite{Euras2012}).
Lineal paths are close for sandstones (with lower $L_2$ values for Sandstone 1 where large pores are occluded with clays or grain fragments),
but for carbonate with much smaller apertures of pores it is way below $L_2$ for sandstones. Also, $L_2$ suggests that computations
with correlations lengths beyond where lineal path goes to zero are not very informative. For very irregular pores on 2D SEM images,
$P$ and $p$ functions look much less similar compared to the ceramics case on \cref{fig:real-xct}. For 2D images it is also visually
easy to analyze full CF maps, as presented on \cref{fig:real-sem-map} for Sandstone 1 sample. Full correlation maps show expanded version of directional functions -
asymmetry is immediately evidenced from $S_2$, $F_{ss}$ and $F_{sv}$ maps. Moreover, the map for $C_2$ shows higher connectivity in y-direction
as compared to x-axis. The interpretation of full correlation maps is less straightforward, but anisotropy patterns in directions different from
those presented in \cref{fig:directions} can be spotted.

\begin{figure*}[pt]
  \centering
  \subfigure[Sandstone 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone1.png}}
  \hfill
  \subfigure[Sandstone 2]{
    \includegraphics[width=0.3\linewidth, frame]{samples/sandstone2.png}}
  \hfill
  \subfigure[Carbonate 1]{
    \includegraphics[width=0.3\linewidth, frame]{samples/carbonate1.png}}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/s2-sem.png}
  \includegraphics[width=0.32\linewidth]{samples/c2-sem.png}
  \includegraphics[width=0.32\linewidth]{samples/l2-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/surf2-sem.png}
  \includegraphics[width=0.32\linewidth]{samples/surfvoid-sem.png}
  \vskip\baselineskip
  \includegraphics[width=0.32\linewidth]{samples/pore-size-sem.png}
  \includegraphics[width=0.32\linewidth]{samples/chord-length-sem.png}
  \caption[]{SEM images and plots of correlation functions for these
    images. Different clusters of the void phase are shown in color.}
  \label{fig:real-sem}
\end{figure*}

\begin{figure*}[pt]
  \centering
  \subfigure[$S_2$]{
    \includegraphics[width=0.45\linewidth]{samples/map-s2.png}}
  \hfill
  \subfigure[$C_2$]{
    \includegraphics[width=0.45\linewidth]{samples/map-c2.png}}
  \vskip\baselineskip
  \subfigure[$F_{ss}$]{
    \includegraphics[width=0.45\linewidth]{samples/map-surf2.png}}
  \hfill
  \subfigure[$F_{sv}$]{
    \includegraphics[width=0.45\linewidth]{samples/map-surfvoid.png}}
  \caption[]{Correlation maps for Sandstone 1.}
  \label{fig:real-sem-map}
\end{figure*}

Finally, we conclude this section with the example of multi-phase soil composite
for which we computed cross-correlation functions \cref{fig:real-multiphase}.
The graph on the right side of the figure shows cross-correlations between
different pairs of phases (normalized by each phase's ratio). One can easily
conclude that organo-mineral complex (phase 2) is mainly located on pore (phase
1) walls. Heavy minerals (phase 4) do not cross-correlate with pores, as they
are mainly lie within the mineral solid phase (phase 3) -- they do show high
cross-correlations. Other CFs can be easily computed for each separate phase,
while $F_{sv}$ can potentially be extended beyond void (or pores) and be
represented by any other phase or their superposition.

@@

\section{Discussion and outlook}
In the light of the end of the previous section it is important to clarify that available
(video) RAM or speed of computations are not the only considerations behind choosing between full correlation map
and directional linear scans. Directional CFs are extremely useful in describing or reconstructing
anisotropic structures \cite{jiao2014chawla,EPL1,tensorPRE}. At the same time, they also much easier to
work with and good candidates for data compression and feature extraction \cite{KarsaninaEJSS}. On the other hand,
full map for $S_2$ allows computing ensemble averaged two-point probability -- in the similar fashion it is
measured with scattering experiments. The usefulness of full correlation maps for other CFs is questionable,
as they are either non-measurable experimentally, e.g., \cite{Gommes_notchord}, or non-applicable for stochastic reconstructions.
The latter refers to our ability to reconstruct almost any structure from its $S_2$ map \cite{fullwood2008}. And while we
have fast and efficient techniques such as phase-recovery to reconstruct the structure from its $S_2$ correlation map, the utilization
of other CFs is complicated \cite{cherkasov2021,cherkasovIC}. Simulated-annealing based methods that operate on the full map are, 
to the best of our knowledge, were not developed to date. This is the reason we did not invest into improvement on existing approaches to compute maps for $L_2$.
